---
title: "Presentation Exercise"

---

## Loading and checking data

The goal is to re-create the FiveThirtyEight figure: **“Baby boomers are the biggest generation in Congress today”**  
(share of members in Congress from each generation, 1919 to 2023).

Portions of the code were initially generated with the assistance of an AI tool and subsequently revised and validated manually.

```{r}
library(tidyverse)
library(lubridate)
library(scales)
```


```{r}
dat <- readr::read_csv("data_aging_congress.csv", show_col_types = FALSE)

glimpse(dat)

```

## Re-create the figure (static)

The original figure is a share-of-members-by-generation plot across Congresses.
We compute the proportion of members in each generation for each Congress.
```{r}
gen_levels <- c(
  "Gilded", "Progressive", "Missionary", "Lost", "Greatest", "Silent",
  "Boomers", "Gen X", "Millennial", "Gen Z"
)

plot_df <- dat %>%
  mutate(
    generation = case_when(
      generation == "baby boomer" ~ "Boomers",
      generation == "Generation X" ~ "Gen X",
      generation == "millennial" ~ "Millennial",
      generation == "Generation Z" ~ "Gen Z",
      TRUE ~ generation
    ),
    generation = factor(generation, levels = gen_levels),
    start_date = as.Date(start_date)
  ) %>%
  filter(!is.na(generation), !is.na(start_date)) %>%
  count(congress, start_date, generation, name = "n") %>%
  group_by(congress, start_date) %>%
  mutate(share = n / sum(n)) %>%
  ungroup()

plot_df %>% arrange(start_date) %>% head()

```

This uses geom_area(position = "fill") to replicate a 100% stacked area chart.
```{r}
p <- ggplot(plot_df, aes(x = start_date, y = share, fill = generation)) +
  geom_area(position = "fill", linewidth = 0) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  labs(
    title = "Baby boomers are the biggest generation in Congress today",
    subtitle = "Share of members in Congress from each generation, 1919 to 2023",
    x = NULL,
    y = NULL,
    fill = NULL,
    caption = "Source: Biographical Directory of the U.S. Congress; VoteView.com (via FiveThirtyEight)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    legend.position = "top",
    legend.justification = "left",
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.caption = element_text(size = 9, color = "gray40"),
    plot.margin = margin(10, 10, 10, 10)
  )

p

```


## Optional: interactive version (toggle generations)


```{r}

library(plotly)


ggplotly(p, tooltip = c("x", "y", "fill"))

```


## Publication-quality table


```{r}
library(gt)


if (!"prop" %in% names(plot_df)) {
  plot_df <- plot_df %>%
    dplyr::group_by(congress, start_date) %>%
    dplyr::mutate(prop = share / sum(share)) %>%
    dplyr::ungroup()
}

tbl_df <- plot_df %>%
  dplyr::select(congress, start_date, generation, prop) %>%
  tidyr::pivot_wider(names_from = generation, values_from = prop) %>%
  dplyr::arrange(dplyr::desc(start_date)) %>%
  dplyr::mutate(
    Year = lubridate::year(start_date),
    Boomers_change_pp = (Boomers - dplyr::lead(Boomers)) * 100
  ) %>%
  dplyr::slice(1:10) %>%
  dplyr::select(Year, congress, Boomers, Boomers_change_pp, `Gen X`, Millennial, `Gen Z`)

```

```{r}
gt(tbl_df) %>%
  tab_header(
    title = md("**Generational Composition of Congress (Most Recent 10 Congresses)**"),
    subtitle = md("Percent shares within each Congress; Boomers column is conditionally highlighted.")
  ) %>%
  tab_spanner(md("**Boomer Trends**"), columns = c(Boomers, Boomers_change_pp)) %>%
  tab_spanner(md("**Younger Generations (Share)**"), columns = c(`Gen X`, Millennial, `Gen Z`)) %>%
  cols_label(
    congress = "Congress",
    Boomers = "Boomers (%)",
    Boomers_change_pp = "Change (pp)",
    `Gen X` = "Gen X (%)",
    Millennial = "Millennial (%)",
    `Gen Z` = "Gen Z (%)"
  ) %>%
  fmt_percent(columns = c(Boomers, `Gen X`, Millennial, `Gen Z`), decimals = 1) %>%
  fmt_number(columns = Boomers_change_pp, decimals = 1) %>%
  data_color(
  columns = Boomers,
  fn = scales::col_numeric(
    palette = c("#f7fbff", "#08306b"),
    domain = range(tbl_df$Boomers, na.rm = TRUE)
   )
  ) %>%
  tab_style(
    style = cell_text(color = "darkgreen", weight = "bold"),
    locations = cells_body(columns = Boomers_change_pp, rows = Boomers_change_pp > 0)
  ) %>%
  tab_style(
    style = cell_text(color = "firebrick", weight = "bold"),
    locations = cells_body(columns = Boomers_change_pp, rows = Boomers_change_pp < 0)
  ) %>%
  tab_source_note(md("**Source:** FiveThirtyEight aging Congress dataset (Biographical Directory; VoteView).")) %>%
  tab_footnote(
    footnote = md("Change is measured in **percentage points** from the prior Congress in this table."),
    locations = cells_column_labels(columns = Boomers_change_pp)
  ) %>%
  cols_align("center")

```
