---
title: "Machine Learning Models Exercise"
---

## Loading packages and data

```{r}
pkgs <- c("tidyverse", "tidymodels", "GGally", "glmnet", "ranger")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org")
  }
}

library(tidyverse)
library(tidymodels)
library(GGally)
library(glmnet)
library(ranger)
```

```{r}
rngseed <- 1234
set.seed(rngseed)


mavo <- readRDS(here::here("ml-models-exercise", "mavo_clean.rds"))

# Recode RACE: combine categories 7 and 88 into 3
mavo <- mavo %>%
  mutate(RACE = as.character(RACE),
         RACE = if_else(RACE %in% c("7", "88"), "3", RACE),
         RACE = factor(RACE))

# Pairwise correlation of continuous predictors
mavo %>% select(Y, DOSE, AGE, WT, HT) %>% ggpairs() + theme_minimal()

# BMI
mavo <- mavo %>% mutate(BMI = WT / (HT^2))
```

## First fits (no CV, no tuning)

```{r}
# Model specifications
lm_spec  <- linear_reg() %>% set_engine("lm")
las_spec <- linear_reg(penalty = 0.1, mixture = 1) %>% set_engine("glmnet")
rf_spec  <- rand_forest(trees = 300) %>%
  set_engine("ranger", seed = rngseed) %>%
  set_mode("regression")

# Fit on full data, all predictors
fit_lm  <- lm_spec  %>% fit(Y ~ ., data = mavo)
fit_las <- las_spec %>% fit(Y ~ ., data = mavo)
fit_rf  <- rf_spec  %>% fit(Y ~ ., data = mavo)

# Predictions and RMSE
pred_all <- bind_rows(
  tibble(model = "Linear", observed = mavo$Y, predicted = predict(fit_lm,  mavo)$.pred),
  tibble(model = "LASSO",  observed = mavo$Y, predicted = predict(fit_las, mavo)$.pred),
  tibble(model = "RF",     observed = mavo$Y, predicted = predict(fit_rf,  mavo)$.pred)
)

pred_all %>%
  group_by(model) %>%
  summarize(rmse = sqrt(mean((observed - predicted)^2)))
```

```{r}
ggplot(pred_all, aes(x = observed, y = predicted, color = model, shape = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ model) +
  labs(title = "Observed vs predicted (no tuning)", x = "Observed Y", y = "Predicted Y") +
  theme_minimal()
```

Linear and LASSO give nearly identical fits. RF fits much closer because it
can capture nonlinearity and interactions.

##  Tuning without CV (apparent resampling)

```{r}
# LASSO penalty grid
las_tune_spec <- linear_reg(penalty = tune(), mixture = 1) %>% set_engine("glmnet")
las_wf <- workflow() %>% add_model(las_tune_spec) %>% add_formula(Y ~ .)
las_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))

las_tune_apparent <- tune_grid(
  las_wf, resamples = apparent(mavo), grid = las_grid, metrics = metric_set(rmse)
)

# autoplot() doesn't work with apparent() (only one resample), so plot manually
collect_metrics(las_tune_apparent, summarize = FALSE) %>%
  ggplot(aes(x = penalty, y = .estimate)) +
  geom_line() + geom_point() +
  scale_x_log10() +
  labs(title = "LASSO tuning (apparent)", y = "RMSE") +
  theme_minimal()
```

```{r}
# Random forest grid: tune mtry and min_n
rf_tune_spec <- rand_forest(trees = 300, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", seed = rngseed) %>%
  set_mode("regression")

rf_wf <- workflow() %>% add_model(rf_tune_spec) %>% add_formula(Y ~ .)
rf_grid <- grid_regular(mtry(range = c(1, 7)), min_n(range = c(1, 21)), levels = 7)

rf_tune_apparent <- tune_grid(
  rf_wf, resamples = apparent(mavo), grid = rf_grid, metrics = metric_set(rmse)
)

collect_metrics(rf_tune_apparent, summarize = FALSE) %>%
  ggplot(aes(x = mtry, y = .estimate, color = factor(min_n))) +
  geom_line() + geom_point() +
  labs(title = "Random forest tuning (apparent)", y = "RMSE", color = "min_n") +
  theme_minimal()
```

LASSO RMSE drops as the penalty shrinks (approaching OLS). RF RMSE drops
with higher `mtry` and lower `min_n` — both signals are misleading because
we are evaluating on training data.

## Tuning with CV

```{r}
set.seed(rngseed)
cv_folds <- vfold_cv(mavo, v = 5, repeats = 5)

# LASSO with CV
las_tune_cv <- tune_grid(
  las_wf, resamples = cv_folds, grid = las_grid, metrics = metric_set(rmse)
)
autoplot(las_tune_cv) + scale_x_log10() + labs(title = "LASSO tuning (CV)")
show_best(las_tune_cv, metric = "rmse", n = 5)
```

```{r}
# Random forest with CV
rf_tune_cv <- tune_grid(
  rf_wf, resamples = cv_folds, grid = rf_grid, metrics = metric_set(rmse)
)
autoplot(rf_tune_cv) + labs(title = "Random forest tuning (CV)")
show_best(rf_tune_cv, metric = "rmse", n = 5)
```


With proper CV both models' RMSEs increase compared to the apparent results, giving an honest estimate of out of sample performance. LASSO often better than the RF, because the underlying signal
is largely linear and RF's earlier advantage was overfitting.
