---
title: "Model Fitting Exercise"
---


## Loading packages and data

```{r}
pkgs <- c("tidyverse", "tidymodels", "here", "skimr", "GGally", "yardstick")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org")
  }
}

library(tidyverse)
library(tidymodels)
library(here)
library(skimr)
library(GGally)
library(yardstick)

# Read the Mavoglurant data
mavo <- read_csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))

# Quick look at the data
glimpse(mavo)
summary(mavo)
```

## Initial exploration: DV over time by DOSE and ID

```{r}
# Plot DV vs TIME, lines per individual (ID), color/facet by DOSE
ggplot(mavo, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ DOSE) +
  labs(
    title = "Drug concentration (DV) over time, stratified by DOSE",
    x = "Time",
    y = "DV (drug concentration)",
    color = "Dose"
  ) +
  theme_minimal()
```

## Data cleaning

```{r}
# keep only OCC == 1
mavo_occ1 <- mavo %>% filter(OCC == 1)

# split into the dosing rows (TIME == 0) and observation rows (TIME != 0)
dose_rows <- mavo_occ1 %>% filter(TIME == 0)
obs_rows  <- mavo_occ1 %>% filter(TIME != 0)

# total drug for each individual = sum of DV across observation rows
total_Y <- obs_rows %>%
  group_by(ID) %>%
  summarize(Y = sum(DV), .groups = "drop")

# get the baseline characteristics from the dosing row
baseline <- dose_rows %>%
  select(ID, DOSE, AGE, SEX, RACE, WT, HT)

# join total Y with baseline
mavo_clean <- baseline %>%
  inner_join(total_Y, by = "ID") %>%
  mutate(
    SEX = factor(SEX),
    RACE = factor(RACE)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

# Save the cleaned data as an RDS file for downstream exercises
saveRDS(mavo_clean, file = "mavo_clean.rds")

glimpse(mavo_clean)
summary(mavo_clean)
```

## Exploratory data analysis

```{r}
# Summary statistics
skim(mavo_clean)

# Distribution of Y
ggplot(mavo_clean, aes(x = Y)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of total drug (Y)", x = "Y", y = "Count") +
  theme_minimal()

# Y vs DOSE (boxplot, treat DOSE as factor for visualization)
ggplot(mavo_clean, aes(x = factor(DOSE), y = Y)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Y by DOSE", x = "DOSE", y = "Y") +
  theme_minimal()

# Y vs AGE
ggplot(mavo_clean, aes(x = AGE, y = Y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Y vs AGE") +
  theme_minimal()

# Y vs WT
ggplot(mavo_clean, aes(x = WT, y = Y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Y vs WT (weight)") +
  theme_minimal()

# Y vs HT
ggplot(mavo_clean, aes(x = HT, y = Y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Y vs HT (height)") +
  theme_minimal()

# Y by SEX
ggplot(mavo_clean, aes(x = SEX, y = Y)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Y by SEX") +
  theme_minimal()

# Pairwise relationships of continuous predictors
mavo_clean %>%
  select(Y, DOSE, AGE, WT, HT) %>%
  GGally::ggpairs()
```

## Model fitting: continuous outcome (Y)

### Model 1: linear regression of Y on DOSE only

```{r}
lin_spec <- linear_reg() %>% set_engine("lm")

# Fit Y ~ DOSE
fit_lm_dose <- lin_spec %>%
  fit(Y ~ DOSE, data = mavo_clean)

tidy(fit_lm_dose)

# Predictions and metrics
pred_dose <- predict(fit_lm_dose, mavo_clean) %>%
  bind_cols(mavo_clean %>% select(Y))

metrics_dose <- pred_dose %>%
  metrics(truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))

metrics_dose
```

### Model 2: linear regression of Y on all predictors

```{r}
fit_lm_all <- lin_spec %>%
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = mavo_clean)

tidy(fit_lm_all)

pred_all <- predict(fit_lm_all, mavo_clean) %>%
  bind_cols(mavo_clean %>% select(Y))

metrics_all <- pred_all %>%
  metrics(truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))

metrics_all
```

### Comparing the two linear models

```{r}
bind_rows(
  metrics_dose %>% mutate(model = "Y ~ DOSE"),
  metrics_all  %>% mutate(model = "Y ~ all predictors")
) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)
```

The model with all predictors generally yields a lower RMSE and higher
R-squared than the DOSE-only model, indicating that including additional
covariates (AGE, SEX, RACE, WT, HT) improves the in-sample fit for the total
drug outcome Y.

## Model fitting: categorical outcome (SEX)

### Model 3: logistic regression of SEX on DOSE only

```{r}
log_spec <- logistic_reg() %>% set_engine("glm")

fit_log_dose <- log_spec %>%
  fit(SEX ~ DOSE, data = mavo_clean)

tidy(fit_log_dose)

# Class predictions and probability predictions
pred_class_dose <- predict(fit_log_dose, mavo_clean, type = "class")
pred_prob_dose  <- predict(fit_log_dose, mavo_clean, type = "prob")

results_log_dose <- mavo_clean %>%
  select(SEX) %>%
  bind_cols(pred_class_dose, pred_prob_dose)

# Accuracy
acc_dose <- accuracy(results_log_dose, truth = SEX, estimate = .pred_class)

# ROC-AUC (use the probability column for the first level of SEX)
prob_col_dose <- paste0(".pred_", levels(mavo_clean$SEX)[1])
auc_dose <- roc_auc(results_log_dose, truth = SEX, !!sym(prob_col_dose))

bind_rows(acc_dose, auc_dose)
```

### Model 4: logistic regression of SEX on all predictors

```{r}
fit_log_all <- log_spec %>%
  fit(SEX ~ Y + DOSE + AGE + RACE + WT + HT, data = mavo_clean)

tidy(fit_log_all)

pred_class_all <- predict(fit_log_all, mavo_clean, type = "class")
pred_prob_all  <- predict(fit_log_all, mavo_clean, type = "prob")

results_log_all <- mavo_clean %>%
  select(SEX) %>%
  bind_cols(pred_class_all, pred_prob_all)

acc_all <- accuracy(results_log_all, truth = SEX, estimate = .pred_class)
prob_col_all <- paste0(".pred_", levels(mavo_clean$SEX)[1])
auc_all <- roc_auc(results_log_all, truth = SEX, !!sym(prob_col_all))

bind_rows(acc_all, auc_all)
```

### Comparing the two logistic models

```{r}
bind_rows(
  bind_rows(acc_dose, auc_dose) %>% mutate(model = "SEX ~ DOSE"),
  bind_rows(acc_all,  auc_all)  %>% mutate(model = "SEX ~ all predictors")
) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)
```

The logistic model that uses all predictors generally has higher accuracy and
ROC-AUC than the model with DOSE alone. DOSE by itself is not very informative
about SEX, while body-size variables such as WT and HT carry stronger signal
for predicting SEX.

# Model Improvement Exercise — Part 1

## Data preparation and train/test split

```{r}
# Drop RACE due to problematic codings; keep Y, DOSE, AGE, SEX, WT, HT
mavo_mod <- mavo_clean %>% select(-RACE)

rngseed <- 1234
set.seed(rngseed)

data_split <- initial_split(mavo_mod, prop = 0.75)
train_data <- training(data_split)
test_data  <- testing(data_split)
```

## Fit two linear models on the training data

```{r}
lin_spec <- linear_reg() %>% set_engine("lm")

fit1_train <- lin_spec %>% fit(Y ~ DOSE, data = train_data)
fit2_train <- lin_spec %>% fit(Y ~ .,    data = train_data)
```

## Assessment 1: training RMSE vs. null model

```{r}
pred_m1_train <- predict(fit1_train, train_data) %>% bind_cols(train_data %>% select(Y))
pred_m2_train <- predict(fit2_train, train_data) %>% bind_cols(train_data %>% select(Y))
null_pred_train <- tibble(.pred = mean(train_data$Y), Y = train_data$Y)

rmse_null <- rmse(null_pred_train, truth = Y, estimate = .pred)
rmse_m1   <- rmse(pred_m1_train,   truth = Y, estimate = .pred)
rmse_m2   <- rmse(pred_m2_train,   truth = Y, estimate = .pred)

bind_rows(
  rmse_null %>% mutate(model = "Null"),
  rmse_m1   %>% mutate(model = "Model 1: Y ~ DOSE"),
  rmse_m2   %>% mutate(model = "Model 2: Y ~ all")
) %>% select(model, .metric, .estimate)
```

## Assessment 2: 10-fold cross-validation

```{r}
set.seed(rngseed)
folds <- vfold_cv(train_data, v = 10)

wf1 <- workflow() %>% add_model(lin_spec) %>% add_formula(Y ~ DOSE)
wf2 <- workflow() %>% add_model(lin_spec) %>% add_formula(Y ~ .)

cv_m1 <- wf1 %>% fit_resamples(resamples = folds, metrics = metric_set(rmse))
cv_m2 <- wf2 %>% fit_resamples(resamples = folds, metrics = metric_set(rmse))

bind_rows(
  collect_metrics(cv_m1) %>% mutate(model = "Model 1: Y ~ DOSE"),
  collect_metrics(cv_m2) %>% mutate(model = "Model 2: Y ~ all")
) %>% select(model, .metric, mean, std_err, n)
```

CV RMSE is higher than in-sample RMSE, especially for Model 2, indicating
some overfitting. Model 2 still outperforms Model 1.

# Model Improvement Exercise — Part 2

## Observed vs predicted for the three models

```{r}
null_pred <- rep(mean(train_data$Y), nrow(train_data))
m1_pred <- predict(fit1_train, train_data)$.pred
m2_pred <- predict(fit2_train, train_data)$.pred

pred_compare <- bind_rows(
  tibble(observed = train_data$Y, predicted = null_pred, model = "Null"),
  tibble(observed = train_data$Y, predicted = m1_pred,   model = "Model 1: Y ~ DOSE"),
  tibble(observed = train_data$Y, predicted = m2_pred,   model = "Model 2: Y ~ all")
)

ggplot(pred_compare, aes(x = observed, y = predicted, color = model, shape = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  labs(title = "Observed vs predicted (training)", x = "Observed Y", y = "Predicted Y") +
  theme_minimal()
```

The null model is one horizontal line; Model 1 forms three horizontal lines
(one per DOSE level); Model 2 spreads along the 45-degree reference.

## Residuals from Model 2

```{r}
resid_df <- tibble(predicted = m2_pred, residual = m2_pred - train_data$Y)

ggplot(resid_df, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(-max(abs(resid_df$residual)), max(abs(resid_df$residual))) +
  labs(title = "Predicted vs residuals (Model 2)", x = "Predicted Y", y = "Residual") +
  theme_minimal()
```

Residuals show negative-skew: larger negative residuals than positive ones,
suggesting some structure the linear model misses.

## Bootstrap uncertainty for Model 2

```{r}
set.seed(rngseed)
boots <- bootstraps(train_data, times = 100)

boot_preds <- map(boots$splits, function(split) {
  fit_b <- lin_spec %>% fit(Y ~ ., data = analysis(split))
  predict(fit_b, train_data)$.pred
})

boot_mat <- do.call(cbind, boot_preds)

boot_summary <- tibble(
  observed = train_data$Y,
  point    = m2_pred,
  median   = apply(boot_mat, 1, median),
  lower    = apply(boot_mat, 1, quantile, probs = 0.055),
  upper    = apply(boot_mat, 1, quantile, probs = 0.945)
)
```

```{r}
ggplot(boot_summary, aes(x = observed)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0, alpha = 0.4, color = "grey60") +
  geom_point(aes(y = point,  color = "Point"),  alpha = 0.8) +
  geom_point(aes(y = median, color = "Median"), alpha = 0.8, shape = 17) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  scale_color_manual(values = c("Point" = "steelblue", "Median" = "firebrick")) +
  labs(title = "Bootstrap predictions, Model 2 (89% CI)",
       x = "Observed Y", y = "Predicted Y", color = "") +
  theme_minimal()
```

# Model Improvement Exercise — Part 3

## Test set evaluation for Model 2

```{r}
test_pred <- predict(fit2_train, test_data)$.pred

combined_pred <- bind_rows(
  tibble(observed = train_data$Y, predicted = m2_pred,   set = "Train"),
  tibble(observed = test_data$Y,  predicted = test_pred, set = "Test")
)

ggplot(combined_pred, aes(x = observed, y = predicted, color = set, shape = set)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  labs(title = "Model 2: train vs test", x = "Observed Y", y = "Predicted Y",
       color = "Dataset", shape = "Dataset") +
  theme_minimal()
```

```{r}
test_metrics <- tibble(.pred = test_pred, Y = test_data$Y) %>%
  rmse(truth = Y, estimate = .pred)

bind_rows(
  rmse_m2      %>% mutate(set = "Train"),
  test_metrics %>% mutate(set = "Test")
) %>% select(set, .metric, .estimate)
```

## Model critique

Both Model 1 and Model 2 beat the null model on training, CV, and test RMSE,
so both extract real signal from the predictors. Model 1 has limited
practical utility because `DOSE` only takes three discrete values, so it can
only output three predictions and cannot distinguish individuals at the same
dose. Model 2 is a meaningful improvement: it has the lowest RMSE everywhere
(train, CV, test), its predictions spread along the 45-degree line, and its
bootstrap medians and 89% intervals are stable. Its main limitations are
asymmetric residuals at the extremes (suggesting nonlinearity the linear
model misses), a small sample (90 train / 30 test), and wider bootstrap
intervals in the tails.

