---
title: "TidyTuesday Exercise: Bird Sightings at Sea (2026-04-14)"
---

TidyTuesday week of 2026-04-14: log-book entries of seabird sightings near
New Zealand, 1969–1990. Data from the Australasian Seabird Mapping Scheme
via Te Papa Tongarewa.

## Setup

```{r}
library(tidyverse)
library(tidymodels)
library(ranger)
library(glmnet)
library(vip)
set.seed(2026)
```

```{r}

data_dir <- if (dir.exists("data")) "data" else "tidytuesday-exercise/data"

birds      <- read_csv(file.path(data_dir, "birds.csv"))
ships      <- read_csv(file.path(data_dir, "ships.csv"))
beaufort   <- read_csv(file.path(data_dir, "beaufort_scale.csv"))
sea_states <- read_csv(file.path(data_dir, "sea_states.csv"))
```

## Data cleaning and joining

Aggregate `birds` to one row per ship record, then join environmental data.

```{r}
# Collapse multiple species rows per ship record into one total count
bird_per_record <- birds |>
  group_by(record_id) |>
  summarise(total_birds = sum(count, na.rm = TRUE), .groups = "drop")

# Join birds + lookup tables to ships; records with no birds get 0
df <- ships |>
  left_join(bird_per_record, by = "record_id") |>
  left_join(beaufort,   by = "wind_speed_class") |>
  left_join(sea_states, by = "sea_state_class") |>
  mutate(
    total_birds = replace_na(total_birds, 0),
    abs_lat     = abs(latitude)   # distance from equator
  )
```

## Exploratory data analysis

```{r}
# Heavy right-skew motivates a log1p outcome
ggplot(df, aes(total_birds + 1)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  scale_x_log10() + theme_minimal()
```

The outcome is heavily right-skewed, so we model `log1p(total_birds)`.

```{r}
# Bird counts by season
df |> filter(!is.na(season)) |>
  ggplot(aes(season, total_birds + 1)) +
  geom_boxplot(fill = "lightgreen") +
  scale_y_log10() + theme_minimal()
```

```{r}
# Spatial pattern shows a clear latitudinal gradient -> use abs_lat.
ggplot(df, aes(longitude, latitude, color = log1p(total_birds))) +
  geom_point(alpha = 0.5, size = 0.6) +
  scale_color_viridis_c() + theme_minimal()
```

## Question

Can environmental conditions and ship activity predict seabird abundance?
Hypothesis: more birds at higher latitudes, in rougher seas, and during
trawling-related activities.

## Preprocessing and split

```{r}
# Outcome + predictors only; drop rows missing the outcome or core predictors.
model_df <- df |>
  transmute(
    log_birds        = log1p(total_birds),
    wind_speed_class = as.integer(wind_speed_class),
    sea_state_class  = as.integer(sea_state_class),
    abs_lat,
    speed            = as.numeric(speed),
    cloud_cover      = factor(cloud_cover),
    precipitation    = factor(precipitation),
    activity         = factor(activity),
    season           = factor(season)
  ) |>
  drop_na(log_birds, wind_speed_class, sea_state_class,
          abs_lat, season, activity)

# 80/20 stratified split + 5-fold CV on the training set.
split <- initial_split(model_df, prop = 0.8, strata = log_birds)
train <- training(split); test <- testing(split)
folds <- vfold_cv(train, v = 5, strata = log_birds)

# Shared recipe: impute, lump rare levels, dummy-encode, normalize.
rec <- recipe(log_birds ~ ., data = train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.02) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

mset <- metric_set(rmse, rsq, mae)  # same metrics for all models
```

## Models

### Linear regression

```{r}
# Interpretable baseline: OLS on the shared recipe
lm_wf <- workflow() |> add_recipe(rec) |>
  add_model(linear_reg() |> set_engine("lm"))
lm_cv <- fit_resamples(lm_wf, folds, metrics = mset)
collect_metrics(lm_cv)
```

### LASSO

```{r}
# Linear with L1 penalty; tune penalty on a log grid 1e-4..1.
lasso_wf <- workflow() |> add_recipe(rec) |>
  add_model(linear_reg(penalty = tune(), mixture = 1) |>
              set_engine("glmnet"))
lasso_tune <- tune_grid(
  lasso_wf, folds,
  grid = grid_regular(penalty(range = c(-4, 0)), levels = 20),
  metrics = mset
)
show_best(lasso_tune, metric = "rmse", n = 3)
```

### Random forest

```{r}
# Captures non-linearities and interactions; tune mtry and min_n.
rf_wf <- workflow() |> add_recipe(rec) |>
  add_model(rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
              set_engine("ranger", importance = "permutation") |>
              set_mode("regression"))
rf_tune <- tune_grid(
  rf_wf, folds,
  grid = grid_regular(mtry(c(2, 8)), min_n(c(2, 20)), levels = 4),
  metrics = mset,
  control = control_grid(save_pred = TRUE)  # keep CV predictions for residual plot
)
show_best(rf_tune, metric = "rmse", n = 3)
```

## CV comparison

```{r}
# Pull the best tuned config from each tuned model for a fair comparison.
best_lasso <- select_best(lasso_tune, metric = "rmse")
best_rf    <- select_best(rf_tune,    metric = "rmse")

cv_results <- bind_rows(
  collect_metrics(lm_cv) |> mutate(model = "Linear"),
  collect_metrics(lasso_tune) |>
    semi_join(best_lasso, by = ".config") |> mutate(model = "LASSO"),
  collect_metrics(rf_tune) |>
    semi_join(best_rf, by = ".config")    |> mutate(model = "Random forest")
) |> select(model, .metric, mean, std_err) |> arrange(.metric, mean)
cv_results
```

```{r}
# CV RMSE with 1-SE error bars.
cv_results |> filter(.metric == "rmse") |>
  ggplot(aes(reorder(model, mean), mean,
             ymin = mean - std_err, ymax = mean + std_err)) +
  geom_col(fill = "steelblue") + geom_errorbar(width = 0.2) +
  labs(x = NULL, y = "CV RMSE") + theme_minimal()
```

```{r}
# Residuals for the best RF: centered on zero indicates no systematic bias.
collect_predictions(rf_tune, parameters = best_rf) |>
  ggplot(aes(.pred, log_birds - .pred)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red") + theme_minimal()
```

## Selection

The random forest has the lowest CV RMSE and highest R², and captures
non-linear effects of latitude, sea state, and ship activity that the
linear models miss. We select it as the final model.

## Test set

```{r}
# Train RF with chosen hyperparameters on full train set; evaluate on test once.
final_fit <- last_fit(finalize_workflow(rf_wf, best_rf),
                      split, metrics = mset)
collect_metrics(final_fit)
```

```{r}
# Predicted vs. observed; closer to the red 45° line is better.
collect_predictions(final_fit) |>
  ggplot(aes(log_birds, .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(color = "red") + theme_minimal()
```

```{r}
# Permutation importance from the final RF.
final_fit |> extract_fit_parsnip() |> vip(num_features = 12)
```

## Discussion

The hypothesis is partly supported. Absolute latitude, Beaufort wind class,
sea state, and ship activity are the strongest predictors, with trawling
records showing clearly higher counts. Wind effects are non-monotonic —
moderate winds yield the highest medians while extremes reduce sightings.
Limitations: heavy zero-inflation, missing weather fields, and
ship of opportunity sampling. A hurdle model and species specific
analyses are natural next steps.
